---
title: "Prediction Assignment"
author: "Ronald Rodríguez - ronaldraxon@gmail.com"
date: "September 09, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(caret)
library(dplyr)
library(ggplot2)
library(utils)
set.seed(301)
trainingUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testingUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainingFile <- "traininig.csv"
testingFile <- "testing.csv" 
```

##1. Overview

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to 
collect a large amount of data about personal activity relatively inexpensively.
These type of devices are part of the quantified self movement - a group of 
enthusiasts who take measurements about themselves regularly to improve their 
health, to find patterns in their behavior, or because they are tech geeks. 
One thing that people regularly do is quantify how much of a particular 
activity they do, but they rarely quantify how well they do it. In this project,
the goal will be to use data from accelerometers on the belt, forearm, arm, and 
dumbell of 6 participants and to predict the manner in which 6 participants 
performed some exercise

##2. Data Loading

create a partition using caret with the training dataset on 70,30 ratio

```{r data downloading, echo = TRUE, eval=FALSE}
download.file(url=trainingUrl,destfile= trainingFile, quiet = TRUE)
download.file(url=testingUrl,destfile= testingFile, quiet = TRUE)
trainingRawData <- read.csv(trainingFile)
testingRawData  <- read.csv(testingFile)

```

##3. Exploratory Analisis
###Selecting variables
```{r select variables, echo = TRUE, eval=FALSE}
summary(trainingRawData)
```
###Cleaning Data
```{r data clean, echo = TRUE, eval=FALSE}
summary(trainingRawData)
```

Anscombe's quartet comprises four datasets that have nearly identical simple descriptive statistics, yet appear very different when graphed. Each dataset consists of eleven (x,y) points. They were constructed in 1973 by the statistician Francis Anscombe to demonstrate both the importance of graphing data before analyzing it and the effect of outliers on statistical properties. He described the article as being intended to counter the impression among statisticians that "numerical calculations are exact, but graphs are rough

```{r training set , echo = TRUE, eval=FALSE}
inTrain  <- createDataPartition(trainingRawData$classe, p=0.7, list=FALSE)
trainSet <- trainingRawData[inTrain, ]
testSet  <- trainingRawData[-inTrain, ]
```

##Appendix

####This code is for the initial set up and required libraries:
```{r, ref.label=c('setup'),echo=TRUE,eval=FALSE}
```




[1]: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
[2]: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
[3]: https://en.wikipedia.org/wiki/Central_limit_theorem
[4]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf